---
title: "Prediction of the delay of a given flight"
output: html_notebook
---

_Aprenentatge Automàtic 1_
_Lucía De Pineda i Adriana Diaz_

```{r}
library(MASS)
library(ggplot2)
library(cclust)
library(dplyr)
library(factoextra)
library(readr)
```

## Data exploration process

Loading the data:
```{r}
Flights = read.csv("airlines.arff", header = FALSE, comment.char = "@")
```

Naming the variables and showing the dataframe:
```{r}
# Variables
names(Flights) = c("Airline", "Flight", "AirportFrom", "AirportTo", "DayOfWeek", "Length", "Time", "Delay")
Flights
```


Analyzing the variables:
```{r}
summary(Flights)
```

Categorical variables: Airline, Flight, AirportFrom, AirportTo, DayOfWeek, Delay

Numeric variables: Time, Length

Frequence of values for target variable:
```{r}
table(Flights$Delay)
hist(Flights$Delay, breaks = 2, main = "Histogram of Delay", col = "lightblue")
```

Histogram and boxplot of length:
```{r}
Length1 = Flights$Length
hist(Length1, main="Histogram of Length (with abnormal data)", col=2)
boxplot(Length1, main="Boxplot of Length (with abnormal data)", col="lightblue")
```


Elimination of abnormal values:
```{r}
Flights = Flights[Flights[,"Length"] >= 300,]
```

Checking there are not missing values:
```{r}
apply(is.na(Flights),2,sum)
```


Transforming variables to factors:
```{r}
Airline=as.factor(Flights$Airline)
Flight=as.factor(Flights$Flight)
AirportFrom=as.factor(Flights$AirportFrom)
AirportTo=as.factor(Flights$AirportTo)
DayOfWeek=as.factor(Flights$DayOfWeek)
Delay=as.factor(Flights$Delay)
Length=Flights$Length
Time=Flights$Time
```

Creation of a new variable:
We convert Time from minutes to hours and Length from miles to km.
```{r}
Speed = (Length / Time) * 96.56064
```

```{r}
length(Airline)
```

Analyzing the distribution of the continuous variables:
```{r}
hist(Time, col=2)
hist(Length, main="Histogram of Length", col=2)
boxplot(Length, main="Boxplot of Length (without abnormal data)", col="lightblue")
```

Applying the logarithm to see if the variables are now Gaussian:
```{r}
hist(log10(Time), col="lightgreen")
hist(log10(Length), col="lightgreen")
```

We apply the logarithm to the variables time and length to normalize them:
```{r}
time_log = log10(Time)
length_log = log10(Length)
```


We combine the transformations into a new dataframe (except Time and Length, we will add them after the visualization of the data):
```{r}
Flights.new = data.frame(Airline,Flight,AirportFrom,AirportTo,DayOfWeek,Length,Time,Delay,Speed)

attach(Flights.new)
```

```{r}
summary(Flights.new)
```


## Visualization of the data

Delay for each day of the week:
```{r}
Delayxday = table(Delay, DayOfWeek)

barplot(Delayxday, beside = TRUE, col = c("pink", "orange"), main = "Delay for each day of the week")

legend("bottomleft", legend=rownames(Delayxday), col="black", 
      fill = c("pink", "orange"), cex=0.65)

```
To plot de delay for each rank of length, we first divide de values of length into different segments:

```{r}
Length.cat = cut(Length, breaks = seq(300, 1439, 200))   
(table(Length.cat))
```
Contingency table:
```{r}
(DelayxLength = table(Delay, Length.cat)) # contingency table

round(prop.table(DelayxLength), digits=3)  # relative freqencies, rounded to 3 digits 
round(prop.table(DelayxLength) * 100, digits=1)  # total percentages

round(prop.table(DelayxLength, 1), digits=3) # table of relative frequencies (row-wise)
(DelayxLength.rel = round(prop.table(DelayxLength, 2), digits=3)) # table of relative frequencies (column-wise)
```
And finally we plot it:
```{r}
DelayxLength = table(Delay, Length.cat)

barplot(DelayxLength, beside = TRUE, col = c("pink", "orange"), main = "Delay per length", cex.names = 0.85)

legend("bottomleft", legend=rownames(DelayxLength), col="black", 
      fill = c("pink", "orange"), cex=0.65)

```
We do the same for Time:
```{r}
Time.cat1 = cut(Time, breaks = seq(25, 425, 50))   
Time.cat2 = cut(Time, breaks = seq(425, 625, 100))   

(table(Time.cat1))
(table(Time.cat2))

Time.cat = c(Time.cat1, Time.cat2)
(table(Time.cat))

Time.cat_eq = cut(Time, breaks = seq(25, 625, 50))
(table(Time.cat_eq))

```

```{r}
(DelayxTime = table(Delay, Time.cat_eq)) # contingency table

round(prop.table(DelayxTime), digits=3)  # relative freqencies, rounded to 3 digits 
round(prop.table(DelayxTime) * 100, digits=1)  # total percentages

round(prop.table(DelayxTime, 1), digits=3) # table of relative frequencies (row-wise)
(DelayxTime.rel = round(prop.table(DelayxTime, 2), digits=3)) # table of relative frequencies (column-wise)
```

```{r}
DelayxTime = table(Delay, Time.cat_eq)

barplot(DelayxTime, beside = TRUE, col = c("pink", "orange"), main = "Delay per time", cex.names = 0.5)

legend("bottomleft", legend=rownames(DelayxTime), col="black", 
      fill = c("pink", "orange"), cex=0.65)

```

Now, we can add the transformated continous variables into the dataframe, also standardizing them:
```{r}
Time = scale(time_log)
Length = scale(length_log)
Speed = scale(Speed)
Flights.new = data.frame(Airline,Flight,AirportFrom,AirportTo,DayOfWeek,Length,Time,Delay,Speed)

attach(Flights.new)
```

# Clustering

We plot our data using two clusters, using the method kmeans.
```{r}
kmeans = kmeans(Flights.new[,c( "Length","Time")], centers=2)
fviz_cluster(kmeans, data=Flights.new[,c(  "Length","Time")])
```

We save our processed data into a csv document, for the next part of the project.
```{r}
write.csv(Flights.new, "airlines2.csv")
```

